---
title: "Problem Set 0"
subtitle: "Bias in Self-reported Turnout"
author: "[YOUR NAME]"
institute: "Villanova University"
date: "Due Date: 2025-01-17 by 11:59PM ET"
output:
  pdf_document: default
  html_document: default
editor_options:
  chunk_output_type: console
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancypagestyle{plain}{\pagestyle{fancy}}
- \fancyhead[L]{}
- \fancyhead[C]{This material should not be shared beyond those who are enrolled in
  this class}
---

Surveys are frequently used to measure political behavior such as voter turnout, but some researchers are concerned about the accuracy of self-reports.  In particular, they worry about possible *social desirability bias* where in post-election surveys, respondents who did not vote in an election lie about not having voted because they may feel that they should have voted.  Is such a bias present in the American National Election Studies (ANES)?  The ANES is a nation-wide survey that has been conducted for every election since 1948.  The ANES conducts face-to-face interviews with a nationally representative sample of adults. The table below displays the names and descriptions
of variables in the `turnout.csv` data file.

--------------------------------------------------------------------------------
 Name                 Description
 -------------------- ----------------------------------------------------------
 `year`               Election year
 
 `ANES`               ANES estimated turnout (percentage)
 
 `VEP`                Voting Eligible Population (in thousands) 
 
 `VAP`                Voting Age Population (in thousands) 
 
 `total`              Total ballots cast for highest office (in thousands) 
 
 `felons`             Total ineligible felons (in thousands) 
 
 `noncitizens`        Total non-citizens (in thousands) 
 
 `overseas`           Total eligible overseas voters (in thousands)
 
 `osvoters`           Total ballots counted by overseas voters (in thousands)
--------------------------------------------------------------------------------


## Question 1

Load the data into R and check the dimensions of the data. Also, obtain a summary of the data.  How many observations are there?  What is the range of years covered in this data set? 

## Answer 1

```{r}
# Data loading differs from original QSS
## Adpating the QSSS code to use rio and download the data directly from github
library("tidyverse")
turnout <- read.csv("../data/turnout.csv")

# if necessary, install qss package
# devtools::install_github("kosukeimai/qss")
# data("turnout", package = "qss")

# alternate methods to load data

# turnout <- import("https://raw.githubusercontent.com/kosukeimai/qss/master/INTRO/turnout.csv")
# write_csv(turnout, "data/import.csv") #saving the downloaded data

dim(turnout)
summary(turnout)
glimpse(turnout)
```

There are `r dim(turnout)[1]` observations in this data set.  The data covers the time period from `r min(turnout$year)` to `r max(turnout$year)`

## Question 2

Calculate the turnout rate based on the voting age population or VAP. Note that for this data set, we must add the total number of eligible overseas voters since the *VAP* variable does not include these individuals in the count. Next, calculate the turnout rate using the voting eligible population or VEP.  What difference do you observe?

## Answer 2

```{r}
## calculate turnout rate based on VAP -- original QSS method
VAPtr <- turnout$total / (turnout$VAP + turnout$overseas) * 100
names(VAPtr) <- turnout$year
VAPtr

## calculate turnout rate based on VEP -- original QSS Method
VEPtr <- turnout$total / turnout$VEP * 100
names(VEPtr) <- turnout$year
VEPtr

## Use mutate instead to add the new variables, turnout rate based on VAP and VEP

turnout <- 
  turnout %>%
  mutate(VAPtr = total/(VAP + overseas) *100,
         VEPtr = total/VEP *100)

## Compare the VEP and VAP turnout rates over time
turnout %>%
  select(year, VAPtr, VEPtr)
```

We find that the VAP turnout rate is consistently lower than the VEP  turnout rate. This result is due to the fact that the VAP includes numerous individuals who are ineligible to vote.

## Question 3

Compute the difference between VAP and ANES estimates of turnout rate. How big is the difference on average?  What is the range of the difference? Conduct the same comparison for the VEP and ANES estimates of voter turnout. Briefly comment on the results.

## Answer 3

```{r}
## compare VAP and ANES -- original QSS Method
diffVAP <- turnout$ANES - VAPtr
mean(diffVAP)
range(diffVAP)

## compare VEP and ANES -- original QSS
diffVEP <- turnout$ANES - VEPtr
mean(diffVEP)
range(diffVEP)

### compare VAP and ANES, VEP and ANES turnout rates with tidyverse
turnout <- 
  turnout %>%
  mutate(diffVAP = ANES - VAPtr,
         diffVEP = ANES - VEPtr)

#generate average and range of differences 
mean(turnout$diffVAP)
range(turnout$diffVAP)
mean(turnout$diffVEP)
range(turnout$diffVEP)

## or, using summarize/summarise:
diff_summary <- 
  turnout %>%
  summarise(diffVAP_mean = mean(diffVAP, na.rm = TRUE),
            diffVAP_min = min(diffVAP),
            diffVAP_max = max(diffVAP),
            diffVEP_mean = mean(diffVEP, na.rm = TRUE),
            diffVEP_min = min(diffVEP),
            diffVEP_max = max(diffVEP))

# even more concisely, with sumarise_at

turnout %>%
  summarise_at(vars(diffVAP, diffVEP),
               funs(mean = mean(., na.rm = TRUE), min, max))

```

On average, the ANES turnout rate is `r round(mean(diffVAP), 2)` percentage points higher than the VAP turnout rate. The difference ranges from `r round(min(diffVAP), 2)` to `r round(max(diffVAP), 2)` percentage points. Comparatively, we find that on average, the ANES turnout rate is `r round(mean(diffVEP), 2)` percentage points higher than the VEP turnout rate. The difference ranges from `r round(min(diffVEP), 2)` to `r round(max(diffVEP), 2)` percentage points. The ANES turnout rate is higher than both the VAP and VEP turnout rates. However, because the VAP turnout rate is consistently lower than the VEP turnout rate, we find that the difference between the VEP and ANES turnout rates is less than the difference between the VAP and ANES turnout rates.

## Question 4

Compare the VEP turnout rate with the ANES turnout rate separately for presidential elections and midterm elections. Note that the data set excludes the year 2006. Does the bias of the ANES vary across election types?

## Answer 4

```{r}
#### Original QSS method
n.obs <- nrow(turnout)
## presidential
pdiffVEP <- diffVEP[c(seq(from = 1, to = n.obs, by = 2), 14)]
## midterm
mdiffVEP <- diffVEP[seq(from = 2, to = n.obs, by = 2)] 
## comparison
summary(pdiffVEP)
summary(mdiffVEP)


#### tidyverse
pres_years <- seq(from = 1980, to = 2008, by=4)

pres_summary <- 
  turnout %>%
  mutate(pres_year = if_else(year %in% pres_years, "yes", "no")) %>%
  group_by(pres_year) %>%
  summarize(diffVEP_mean = mean(diffVEP, na.rm = TRUE),
            diffVEP_min  = min(diffVEP),
            diffVEP_max  = max(diffVEP))

pres_summary
```

On average, the bias is higher among presidential elections than midterm elections. The ANES turnout rate is on average `r round(pres_summary[2, 2], 2)` percentage points higher than the VEP turnout rate in presidential elections.  In midterm elections, the ANES turnout rate is on average `r round(pres_summary[1, 2], 2)` percentage points higher than the VEP turnout rate.

## Question 5

Divide the data into half by election years such that you subset the data into two periods.  Calculate the difference between the VEP turnout rate and the ANES turnout rate separately for each period. Has the bias of the ANES increased over time?

## Answer 5

```{r}
### Original QSS method
earlier.diffVEP <- diffVEP[1:(n.obs/2)] # first half
later.diffVEP <- diffVEP[(n.obs/2 + 1):n.obs] # second half
## comparison
summary(earlier.diffVEP)
summary(later.diffVEP)

### tidyverse
first_half <- seq(from = 1980, to = 1992, by = 2)

turnout <- 
  turnout %>%
  mutate(first_half = if_else(year %in% first_half, "yes", "no"))

half_summary <-
  turnout %>%
  group_by(first_half) %>%
  summarize(diffVEP_mean = mean(diffVEP, na.rm = TRUE),
            diffVEP_min  = min(diffVEP),
            diffVEP_max  = max(diffVEP))

half_summary
```

It is worth noting that the median differences between the ANES and VEP turnout rates are very similar between the two time periods. For the first half of election years, the median difference is `r round(median(earlier.diffVEP), 2)` percentage points while for the second half the median difference is `r round(median(later.diffVEP), 2)` percentage points. However, on average, we find that the bias of the ANES has increased over time. For the first half of election years, the ANES turnout rate is on average `r round(mean(earlier.diffVEP), 2)` percentage points higher than the VEP turnout rate. Comparatively, the ANES is on average `r round(mean(later.diffVEP), 2)` percentage points higher than the VEP turnout rate among the later election years.

## Question 6

The ANES does not interview overseas voters and prisoners. Calculate an adjustment to the 2008 VAP turnout rate. Begin by subtracting the total number of ineligible felons and non-citizens from the VAP to calculate an adjusted VAP. Next, calculate an adjusted VAP turnout rate, taking care to subtract the number of overseas ballots counted from the total ballots in 2008. Compare the adjusted VAP turnout with the unadjusted VAP, VEP, and the ANES turnout rate. Briefly discuss the results.
  
## Answer 6

```{r}
## Original QSS
adj.VAP <- turnout$VAP[14] - (turnout$felons[14] + turnout$noncit[14])
adj.VAPtr <- (turnout$total[14] - turnout$osvoters[14]) / adj.VAP * 100
adj.diff <- turnout$ANES[14] - adj.VAPtr
## compare adjusted difference to non-adjusted VAP as well as VEP
adj.diff 
diffVAP[14] 
diffVEP[14]

### tidyverse
turnout_2008 <- 
  turnout %>%
  filter(year == 2008) %>%
  mutate(adj.VAP = VAP - felons - noncit,
         adj.VAPtr = (total - osvoters) / adj.VAP * 100,
         adj.diffVAP =  ANES - adj.VAPtr)

#Display relevant variables for comparison
select(turnout_2008, adj.VAPtr, VAPtr, VEPtr, ANES, 
       adj.diffVAP, diffVAP, diffVEP)
```

The adjusted VAP turnout rate is closer to the ANES turnout rate than the non-adjusted VAP or VEP turnout rates of 2008. The adjusted VAP turnout rate is `r round(adj.diff, 2)` percentage points lower than the ANES turnout rate. The non-adjusted VAP turnout rate is `r round(diffVAP[14], 2)` percentage points lower than the ANES turnout rate while the VEP turnout rate is `r round(diffVEP[14], 2)` percentage points lower.  In general it appears that the ANES overestimates turnout rates. However, by adjusting the VAP rate to reflect the conditions of the ANES survey, we reduce the difference in estimated turnout rates.
