---
title: "PSC4375: Summarizing bivariate relationships: cross-tabs, scatterplots, and correlation"
subtitle: "Week 4: Lecture 8"
author: "Prof. Weldzius"
institute: "Villanova University"
date: "Slides Updated: `r Sys.Date()`"
output:
  beamer_presentation:
    theme: "CambridgeUS"
    colortheme: "seahorse"
    fonttheme: "structurebold"
---

```{css,echo = F}
.small .remark-code { /*Change made here*/
  font-size: 85% !important;
}
.tiny .remark-code { /*Change made here*/
  font-size: 50% !important;
}
```

```{r,include=F}
set.seed(123)
options(width=60)
knitr::opts_chunk$set(fig.align='center',fig.width=9,fig.height=5,message=F,warning=F)
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
```

# Effect of assasination attempts 

\pause 
\footnotesize
```{r}
library(tidyverse)
data(leaders, package = "qss")
head(leaders[,1:7])
```


# Before we begin with lesson, Pset 2 help

```{r,  fig.show='hide'}
PolityAfter <- leaders %>%
  group_by(civilwarbefore) %>%
  summarize(polityafter_mean = mean(polityafter))

PolityAfterPlot <- PolityAfter %>%
  ggplot(aes(x=as.factor(civilwarbefore), y = polityafter_mean)) +
  geom_bar(stat = "identity", alpha=0.75) +
  scale_x_discrete(labels = c("No civil war","Civil war")) +
  labs(title = "Dem. score after assasin. attempt",
       y = "Average Democracy Score", x = "") +
  theme_bw()

```

# Before we begin with lesson, Pset 2 help

```{r,fig.width=3.8, fig.height=2.8}
PolityAfterPlot
```

# More Pset 2 help!

Question 5 update:

 - Given that the number of children might be a confounder for the relationship between number of girls and voting, let’s estimate the effects using statistical control for the number of children using the following steps:

    - Create one subset of the data called `girls_123` that restricts to judges with one, two or three children and have at least one girl.
    - Create another subset of the data called `nogirls_123` that restricts to judges with one, two or three children and have no girls.
    - Calculate the mean of `progressive_vote` within levels of the numbers of kids (`num_kids`) for each of these subsets and save these vectors as `girls_vote_by_nkids` and `nogirls_vote_by_nkids.`
    - Use `inner_join` to combine the two data subsets then estimate the average treatment effect within levels, saving this vector as `ate_nkids`.

# More Pset 2 help!

  - Use `inner_join` to combine the two data subsets...

\small 
```{r}
PolityAfter <- leaders %>%
  group_by(civilwarbefore) %>%
  summarize(polityafter_mean = mean(polityafter))

PolityBefore <- leaders %>%
  group_by(civilwarbefore) %>%
  summarize(politybefore_mean = mean(politybefore))

PolityCombine <- inner_join(PolityAfter,PolityBefore)
PolityCombine
```


# Contingency tables

\pause 

  - With two categorical variables, we can create **contingency tables** \pause 
    - Also known as **cross-tabs** \pause 
    - Rows are the values of one variable, columns the other \pause 

\small     
```{r}
leaders %>%
  group_by(civilwarbefore,civilwarafter) %>%
  count() %>%
  spread(civilwarafter, n)
```
\normalsize 

  - Quick summary how the two variables "go together" 

---

# Cross-tabs with proportions

```{r}
leaders %>%
  group_by(civilwarbefore,civilwarafter) %>%
  count() %>%
  ungroup() %>%
  mutate(prop = n/ sum(n)) %>%
  select(-n) %>%
  spread(civilwarafter, prop, drop = T)
```

---

# Cross-tabs with proportions (by row)

```{r}
leaders %>%
  group_by(civilwarbefore,civilwarafter) %>%
  count() %>%
  ungroup() %>%
  group_by(civilwarbefore) %>%
  mutate(prop = n/ sum(n)) %>%
  select(-n) %>%
  spread(civilwarafter, prop, drop = T)
```

---

# Scatterplot

```{r, echo=F, fig.width=4, fig.height=3}
leaders %>%
  ggplot(aes(x = politybefore, y = polityafter)) +
  geom_point(shape = 21) +
  labs(title = "Democracy before/after assasination attempts",
       x = "Democracy Level (Before)",
       y = "Democracy Level (After)") +
  theme_bw() +
  theme(plot.title = element_text(size=12))
```

---

# Scatterplot

  - Each point on the scatterplot $(x_i, y_i)$
  - Use `geom_point()` function in `ggplot`
  
```{r, echo=T, fig.show='hide'}
leaders %>%
  ggplot(aes(x = politybefore, y = polityafter)) +
  geom_point(shape = 21) +
  labs(title = "Democracy before/after assasination attempts",
       x = "Democracy Level (Before)",
       y = "Democracy Level (After)") +
  theme_bw() +
  theme(plot.title = element_text(size=12))
```

---

# Scatterplot

\footnotesize
```{r, echo=T}
leaders[1, c("politybefore","polityafter")]
```


```{r, echo=F, fig.width=3, fig.height=2}
leaders %>%
  ggplot(aes(x = politybefore, y = polityafter)) +
  geom_point(shape = 21, alpha=0) +
  geom_point(data = leaders[1, ], aes(x = politybefore, y = polityafter), color = "red", shape = 21) +
  labs(title = "Democracy before/after assasination attempts",
       x = "Democracy Level (Before)",
       y = "Democracy Level (After)") +
  theme_bw() +
  theme(plot.title = element_text(size=8),
        axis.title.x = element_text(size=7),
        axis.title.y = element_text(size=7))
```

---

# Scatterplot

\footnotesize
```{r, echo=T}
leaders[2, c("politybefore","polityafter")]
```


```{r, echo=F, fig.width=3, fig.height=2}
leaders %>%
  ggplot(aes(x = politybefore, y = polityafter)) +
  geom_point(shape = 21, alpha=0) +
  geom_point(data = leaders[1, ], aes(x = politybefore, y = polityafter), shape = 21) +
    geom_point(data = leaders[2, ], aes(x = politybefore, y = polityafter), color = "red", shape = 21) +
  labs(title = "Democracy before/after assasination attempts",
       x = "Democracy Level (Before)",
       y = "Democracy Level (After)") +
  theme_bw() +
  theme(plot.title = element_text(size=8),
        axis.title.x = element_text(size=7),
        axis.title.y = element_text(size=7))
```

---

# Scatterplot

\footnotesize
```{r, echo=T}
leaders[3, c("politybefore","polityafter")]
```


```{r, echo=F, fig.width=3, fig.height=2}
leaders %>%
  ggplot(aes(x = politybefore, y = polityafter)) +
  geom_point(shape = 21, alpha=0) +
  geom_point(data = leaders[1, ], aes(x = politybefore, y = polityafter), shape = 21) +
  geom_point(data = leaders[2, ], aes(x = politybefore, y = polityafter), shape = 21) +
  geom_point(data = leaders[3, ], aes(x = politybefore, y = polityafter), color = "red", shape = 21) +
  labs(title = "Democracy before/after assasination attempts",
       x = "Democracy Level (Before)",
       y = "Democracy Level (After)") +
  theme_bw() +
  theme(plot.title = element_text(size=8),
        axis.title.x = element_text(size=7),
        axis.title.y = element_text(size=7))
```


---

# Scatterplot

\footnotesize
```{r, echo=T}
leaders[3, c("politybefore","polityafter")]
```


```{r, echo=F, fig.width=3, fig.height=2}
leaders %>%
  ggplot(aes(x = politybefore, y = polityafter)) +
  geom_point(shape = 21) +
  labs(title = "Democracy before/after assasination attempts",
       x = "Democracy Level (Before)",
       y = "Democracy Level (After)") +
  theme_bw() +
  theme(plot.title = element_text(size=8),
        axis.title.x = element_text(size=7),
        axis.title.y = element_text(size=7))
```

---

# How big is big?

\pause 
  - Would be nice to have a standard summary of how similar variables are \pause
    - Problem: variables on different scales! \pause
    - Needs a way to put any variable on common units \pause
    
  - **z-score** to the rescue! \pause
  
  $$
  \text{z-score of } x_i = \frac{x_i - \text{mean of } x}{\text{standard deviation of }x}
  $$
 \pause
 
  - Crucial property: z-scores don't depend on units \pause
  
  $$
  \text{z-score of } (ax_i + b) = \text{z-score of } x_i
  $$
  
---

# Correlation
 \pause
 
  - How do variables move together on average? \pause
  - When $x_i$ is big, what is $y_i$ likely to be? \pause
    - Positive correlation: when $x_i$ is big, $y_i$ is also big \pause
    - Negative correlation: when $x_i$ is big, $y_i$ is small \pause
    - High magnitude of correlation: data cluster tightly around a line \pause
    
  - The technical definition of the **correlation coefficient**: \pause
  
  $$
  \frac{1}{n-1} \sum_{i=1}^{n}{\left[ (\text{z-score for }x_i)\times(\text{z-score for }y_i) \right]}
  $$
  
---

# Correlation intuition:
 
\begin{center}
\includegraphics[width=.75\textwidth]{figs/corr1.png}
\end{center}

---

# Correlation intuition:

 
\begin{center}
\includegraphics[width=.75\textwidth]{figs/corr2.png}
\end{center}
\pause

  - Large values of $X$ tend to occur with large values of $Y$ \pause
    - (z-score for $x_i$) $\times$ (z-score for $y_1$) = (pos. num.) $\times$ (pos. num.) $= +$ \pause
  - Small values of $X$ tend to occur with small values of $Y$   \pause
    - (z-score for $x_i$) $\times$ (z-score for $y_1$) = (neg. num.) $\times$ (neg. num.) $= +$ \pause
  - If these dominate $\rightsquigarrow$ positive correlation
  
---

# Correlation intuition:

 
\begin{center}
\includegraphics[width=.75\textwidth]{figs/corr3.png}
\end{center}
\pause

  - Large values of $X$ tend to occur with small values of $Y$ \pause
    - (z-score for $x_i$) $\times$ (z-score for $y_1$) = (pos. num.) $\times$ (neg. num.) $= -$ \pause
  - Small values of $X$ tend to occur with large values of $Y$   \pause
    - (z-score for $x_i$) $\times$ (z-score for $y_1$) = (neg. num.) $\times$ (pos. num.) $= -$ \pause
  - If these dominate $\rightsquigarrow$ negative correlation  
    
---

# Properties of correlation coefficient

 \pause
 
  - Correlation measures **linear** association. \pause
  - Interpretation: \pause
    - Correlation is between -1 and 1 \pause
    - Correlation of 0 means no linear association \pause
    - Positive correlations $\rightsquigarrow$ positive associations \pause
    - Negative correlations $\rightsquigarrow$ negative associations \pause
    - Closer to -1 or 1 means stronger association \pause
  - Order doesn't matter: `cor`(x,y) = `cor`(y,x) \pause
  - Not affected by changes of scale: \pause
    - `cor`(x,y) = `cor`(ax+b, cy+d)
    - Celsius vs. Fahrenheit; dollars vs. pesos; cm vs. in.
    
---

# Correlation in R

  - Use the `cor()` function \pause 
  
```{r}
leaders %>%
  select(politybefore, polityafter) %>%
  cor()
```
  
  - Very highly correlated!
  