---
title: "PSC4375: Visualizing Distributions"
subtitle: "Week 3: Lecture 5"
author: "Prof. Weldzius"
institute: "Villanova University"
date: "Slides Updated: `r Sys.Date()`"
output:
  beamer_presentation:
    theme: "CambridgeUS"
    colortheme: "seahorse"
    fonttheme: "structurebold"
---

```{css,echo = F}
.small .remark-code { /*Change made here*/
  font-size: 85% !important;
}
.tiny .remark-code { /*Change made here*/
  font-size: 50% !important;
}
```

```{r,include=F}
set.seed(123)
options(width=60)
knitr::opts_chunk$set(fig.align='center',fig.width=9,fig.height=5,message=F,warning=F)
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
```

# Studying political efficacy

  - 2002 WHO survey of people in China and Mexico
  - Goal: determine feelings of political efficacy
  - Question: "How much say do you have in getting the government to address issues that interest you?"
    1. No say at all
    2. little say
    3. some say
    4. a lot of say
    5. unlimited say
    
---

# Data

 - Load the data:
 
```{r}
library(tidyverse)
data(vignettes, package = "qss")
head(vignettes)
## Also works if you downloaded the data:
# vignettes <- read.csv("data/vignettes.csv")
```

---

# Contingency table

  - `count()` shows how many units are in each category of a variable:
```{r}
vignettes %>%
  count(self)
```

---

# Contingency table (continued)

  - using `group_by` we can convert these counts into proportions of units:
  
```{r}
vignettes %>%
  group_by(self) %>%
  count() %>%
  ungroup() %>%
  mutate(prop = n / sum(n))
  
```

---

# Useful way to visualize this information: **barplot**

```{r,echo=F,fig.width=4, fig.height=3}
ggplot(data = vignettes,
       aes(x = as.factor(self))) +
  geom_bar(aes(y=..prop..,
               group = 1)) + # add a barplot layer
  scale_x_discrete(labels = c("None", "A little",
                  "Some", "A lot", "Unlimited")) +
  xlab("Self-reported political efficacy") +
  ylab("Proportion of Respondents")
```

---

# Barplots in R

  - The `barplot()` function can help us visualize a categorical variable:
  
```{r, echo=T, fig.show='hide'}
vignettes %>%
  ggplot(aes(x = as.factor(self),
             y=..prop..,
             group = 1)) +
  geom_bar() + 
  scale_x_discrete(labels = c("None", "A little",
                  "Some", "A lot", "Unlimited")) +
  xlab("Self-reported political efficacy") +
  ylab("Proportion of Respondents")
```

  - Arguments:
    - `aes()`: the aestetic mapping of the plot (what you see)
    - `scale_x_discrete()`: changes the scale of the axis
    - `xlab()`, `ylab()` are axis labels
    
---

# Histogram

  - **Histograms** visualize density of a continuous/numeric variable
  
```{r,echo=F,fig.width=3.2, fig.height=2.7}
vignettes %>%
  ggplot(aes(x = age, y = ..density..)) +
  geom_histogram(binwidth = 5, # how wide for each bin
                 boundary = 0, # bin position
                 alpha=0.75) + # reduces opaqueness
  scale_x_continuous(breaks = seq(20, 80,by = 10)) +
  labs(title = "Distribution of Respondent's Age",
       y = "Density",
       x = "Age") + 
  theme_bw()
```

---

# How to create histograms?

  - How to create a histogram by hand:
    1. create bins along the variable of interest
    2. count number of observations in each bin
    3. **density** = bin height
    
  $$
  \text{density} = \frac{\text{proportion of observations in bin}}{\text{bin width}}
  $$
  
  - The areas of the bins $=$ proportion of observations in those bins.
    - $\rightsquigarrow$ area of the blocks sum to 1 (100%)
    - Can lead to confusion: height of block can go above 1!
    - With equal-width bins, height is proportional to proportion in bin.
    
---

# Histograms in R

  - In ggplot, we use `geom_histogram()`:

```{r, echo=T, fig.show='hide'}
vignettes %>%
  ggplot(aes(x = age,
             y = ..density..)) +
  geom_histogram(binwidth = 5, # how wide for each bin
                 boundary = 0, # bin position
                 alpha=0.75) + # reduces opaqueness
  scale_x_continuous(breaks = seq(20, 80,by = 10)) +
  labs(title = "Distribution of Respondent's Age",
       y = "Density",
       x = "Age") + 
  theme_bw()
```

  - Other arguments:
    -`labs` sets the titles for the plot (used `xlab` and `ylab` in previous plot)
    -`scale_x_continuous` sets the scale for the x-axis
    
    
---

# Histograms in R: adding a vertical `median` line

```{r,echo=F,,fig.width=4, fig.height=3}
vignettes %>%
  ggplot(aes(x = age,
             y = ..density..)) +
  geom_histogram(binwidth = 5, # how wide for each bin
                 boundary = 0, # bin position
                 alpha=0.75) + # reduces opaqueness
  geom_vline(xintercept = median(vignettes$age)) + 
  annotate(geom = "text", x = median(vignettes$age),
           y=.03, label = "Median", hjust = -0.1) +
  scale_x_continuous(breaks = seq(20, 80,by = 10)) +
  labs(title = "Distribution of Respondent's Age",
       y = "Density",
       x = "Age") + 
  theme_bw()
```

    
---

# Histograms in R: adding a vertical `median` line

```{r,echo=T, results='hide'}
vignettes %>%
  ggplot(aes(x = age,
             y = ..density..)) +
  geom_histogram(binwidth = 5, # how wide for each bin
                 boundary = 0, # bin position
                 alpha=0.75) + # reduces opaqueness
  geom_vline(xintercept = median(vignettes$age)) + 
  annotate(geom = "text", x = median(vignettes$age),
           y=.03, label = "Median", hjust = -0.1) +
  scale_x_continuous(breaks = seq(20, 80,by = 10)) +
  labs(title = "Distribution of Respondent's Age",
       y = "Density",
       x = "Age") + 
  theme_bw()
```

---

# Boxplot

  - A **boxplot** can characterize the distribution of continuous variables

```{r, echo=F}
vignettes %>%
  ggplot(aes(y = age)) +
  geom_boxplot()
```

