---
title: "PSC4375: Descriptive Statistics"
subtitle: "Week 2: Lecture 4"
author: "Prof. Weldzius"
institute: "Villanova University"
date: "Slides Updated: `r Sys.Date()`"
output:
  beamer_presentation:
    theme: "AnnArbor"
    colortheme: "dolphin"
    fonttheme: "structurebold"
  classoption: t
---

```{css,echo = F}
.small .remark-code { /*Change made here*/
  font-size: 85% !important;
}
.tiny .remark-code { /*Change made here*/
  font-size: 50% !important;
}
```

```{r,include=F}
set.seed(123)
options(width=60)
knitr::opts_chunk$set(fig.align='center',fig.width=9,fig.height=5,message=F,warning=F)
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
```

# Lots of data

  * Data from study of the effect of minimum wage
  
```{r}

```
  
---

# Lots and lots of data

  
```{r}
# head(minwagewageAfter, n = 200)
```

---

# How to summarize data

  * How should we summarize the wages data? Many possibilities!
    - Up to now: focus on **averages** or means of variables
    
  * Two salient features of a variable that we want to know:
    - **Central tendency**: where is the middle/typical/average value
    - **Spread** around the center: are all values to the center or spread out?
    
---

# Center of the data

  * "Center" of the data: typical/average value
  * **Mean**: sum of the values divided by the number of observations
  $$
  \bar{x} = \frac{1}{n} \sum_{i=1}^{n}{x_i}
  $$
  * **Median**:
  $$
  \text{median} = 
   \begin{cases}
      \text{middle value} & \text{if number of entries is odd} \\
      \frac{\text{sum of two middle values}}{2} & \text{if number of entries is even}
    \end{cases}  
  $$
  * In **R**: `mean()` and `median()` 
  
---

# Mean vs median

  * Median more robust to **outliers**:
    - Example 1: data = ${0,1,2,3,5}$. Mean? Median?
    \vspace{10pt}
    - Example 2: data = ${0,1,2,3,100}$. Mean? Median?
    \vspace{10pt}
  * What does Mark Zuckerberg do to the mean vs. median income?  
  
---

# Spread of the data

  * Are the values of the variable close to the center?
  * **Range**: $[\text{min}(X), \text{max}(X)]$
  * **Quantile** (quartile, percentile, etc.): divide data into equal sized groups.
    - 25th percentile: lower quartile (25% of the data below this value)
    - 50th percentile: median (50% of the data below this value)
    - 75th percentile: upper quartile (75% of the data below this value)
  * **Interquartile range** (IQR): a measure of variability
    - How spread out is the middle half of the data?
    - Is most of the data really close to the median or are the values spread out?
  * **R** function: `range()`, `summary()`, `IQR()`
  
---

# Standard deviation

  * **Standard deviation**: On average, how far away are data points from the mean?
  
  $$
  \text{standard deviation} = \sqrt{\frac{1}{n-1}\sum_{i=1}^{n}{\left( x_i - \bar{x} \right)^2}}
  $$
  * Steps:
    1. Subtract each data point by the mean
    2. Square each resulting difference
    3. Take the sum of these values
    4. Divide by $n-1$ (or $n$, doesn't matter much)
    5. Take the square root
  * **Variance**: standard deviation$^2$
  * Why not just take the average deviations from mean without squaring? 
  
---

# How large is large?

  * Is a wage of 5.30 an hour large?
  * Better question: is 5.30 large relative to the distribution of the data?
    - Big in one dataset might be small in another!
    - Different units, difference spreads of the data, etc.
  * Need a way to put any variable on **common units**
  * **z-score**: 
  
  $$
  \text{z-score of }x_i = \frac{x_i - \text{mean of }x}{\text{standard deviation of }x} 
  $$
  * Interpretation:
    - Positive values above the mean, negative values below the mean
    - Units now on the scale of **standard deviations away from the mean**
    - Intuition: data more than 3 SDs away from mean are rare

---

# z-score example

  * Jane works at The Grog where there's a tip jar.
  * She's been keeping track of herh daily tips:
    - Average tip of $1.56 with a standard deviation of 20 cents. 
  * Yesterday, Jane got $1.86 in tips. How big is this?
  \vspace{10pt}
  * Today she got $0.56, what about that? 
  
    