---
title: "PSC4375: Loops & Predicting"
subtitle: "Week 5: Lecture 9"
author: "Prof. Weldzius"
institute: "Villanova University"
date: "Slides Updated: `r Sys.Date()`"
output:
  beamer_presentation:
    theme: "CambridgeUS"
    colortheme: "seahorse"
    fonttheme: "structurebold"
---

```{css,echo = F}
.small .remark-code { /*Change made here*/
  font-size: 85% !important;
}
.tiny .remark-code { /*Change made here*/
  font-size: 50% !important;
}
```

```{r,include=F}
set.seed(123)
options(width=60)
knitr::opts_chunk$set(fig.align='center',fig.width=9,fig.height=5,message=F,warning=F)
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
```

# 2016 US Presidential Election

  \begin{center}
  \includegraphics[width=.5\textwidth]{figs/2016election.png}
  \end{center} \pause 
  
  - 2016 election popular vote: \pause
    - Clinton: 65,853,516 (48.2%)
    - Trump: 62,984,825 (46.1%) \pause
  - Why did Trump win? **Electoral college**
    - Trump: 304, Clinton: 227 \pause
  - Election determined by 77,744 votes (margins in WI, MI, PA)
    - 0.056% of the electorate (~136million)
    
---

# Predicting US Presidential Elections

 \pause
 
  - **Electoral college system** \pause
    - Must win an absolute majority of 538 electoral votes \pause
    - 538 = 435 (House of Representatives) + 100 (Senators) + 3 (DC) \pause
    - Must win at least 270 votes \pause
    - nobody wins an absolute majority $\rightsquigarrow$ House vote \pause
  - Must predict winner of each state \pause
  
  \begin{center}
  \includegraphics[width=.6\textwidth]{figs/2016electcol.png}
  \end{center}  

---

# Prediction strategy

 \pause
 
  - Predict state-level support for each candidate using polls \pause
  - Allocate electoral college votes of that state to its predicted winner \pause
  - Aggregate EC votes across states to determine the predicted winner \pause
  - Coding strategy: \pause
  
    1. For each state, subset to polls within that state \pause
    2. Further subset the latest polls \pause
    3. Average the latest polls to estimate support for each candidate \pause
    4. Allocate the electoral votes to the candidate who has greatest support \pause
    5. Repeat this for all states and aggregate the electoral votes \pause 
    
  - Sounds like a lot of subsets, ugh...
  
---

# Multiplication {.t}

 \pause
 
  - Let's create a new variable that multiples each value in a vector by 2: \pause
    - Easy in R: `values * 2` \pause
    - Pretend you didn't know this approach

```{r, echo=F}
require(tidyverse)
```

---

# Multiplication {.t}

\footnotesize  
```{r, eval=F}
values <- c(2,4,6)
```


---

# Multiplication {.t}

\footnotesize  
```{r, eval=F}
values <- c(2,4,6)

## number of values
n <- length(values)
```

# Multiplication {.t}

\footnotesize  
```{r, eval=F}
values <- c(2,4,6)

## number of values
n <- length(values)

## create container to hold results
results <- rep(NA, times = n)
```

# Multiplication {.t}

\footnotesize  
```{r}
values <- c(2,4,6)

## number of values
n <- length(values)

## create container to hold results
results <- rep(NA, times = n)

## multiply each value by 2
results[1] <- values[1] * 2
results[2] <- values[2] * 2
results[3] <- values[3] * 2
```

---

# Multiplication {.t}

\footnotesize  
```{r}
values <- c(2,4,6)

## number of values
n <- length(values)

## create container to hold results
results <- rep(NA, times = n)

## multiply each value by 2
results[1] <- values[1] * 2
results[2] <- values[2] * 2
results[3] <- values[3] * 2

results
```


---

# Loops in R

 \pause
 
  - What if you had more values? Not efficient! \pause
    - **for loop**: a way to iteratively execute the same code multiple times. \pause
  - Basic structure: \pause

\small  
```{r, echo=T, eval=FALSE}
for (i in X) {
  expression1
  expression2
  ...
  expression3
}
```

 \pause
 
  - Elements of a loop:  \pause
    - `i`: counter (can use any name)  \pause
    - `X`: vector containing a set of ordered values the counter takes \pause
    - `expression`: a set of expressions that will be repeatedly evaluated. \pause
    - `{ }`: curly braces to define beginning and end of the loop. \pause
  - Indentation is important for readability of the code.
  
---

# Loop example:

\small 
```{r}
values <- c(2,4,6)
n <- length(values)
results <- rep(NA, times = n)

## begin loop
for (i in 1:n) {
  results[i] <- values[i] * 2
  print(str_c(values[i], " times 2 is equal to ", results[i]))
}
```

---

# 2016 polling prediction

  - Election data: `pres.csv`

\footnotesize   
| Name           | Description |
|----------------|:------------------------------------------------------------------------------------------|
| `state_abb`         |      abbreviated name of state |
| `clinton` | Clinton's vote share (percentage) |
| `trump` | Trump's vote share (percentage) |

  - Polling data `polls16.csv`

\footnotesize   
| Name           | Description |
|----------------|:------------------------------------------------------------------------------------------|
| `state`         |      abbreviated name of state in which polls was conducted |
| `middate` | middate of the period when polls was conducted |
| `daysleft` | number of days between middate and election day  |
| `pollster` | name of organization conducting poll |
| `clinton` | predicted support for Clinton (percentage) |
| `trump` | predicted support for Trump (percentage) |

---

# Some preprocessing

```{r}
## download; don't forget to setwd()
pres16 <- read_csv("../data/pres2016.csv")
polls16 <- read_csv("../data/polls2016.csv")

## calculate Trump's margin of victory
polls16 <- polls16 %>%
  mutate(margin = Trump - Clinton)
pres16 <- pres16 %>%
  mutate(margin = Trump - Clinton)
```

---

# What does the data look like?

\footnotesize
```{r}
head(polls16)
```

---

# Poll prediction for each state {.t}

\footnotesize
```{r}
## place holder
poll.pred <- rep(NA, 51)

## get list of unique state names to iterate over
state_names <- unique(polls16$state)

## add labels to place holder
names(poll.pred) <- state_names
```
---

# Poll prediction for each state {.t}

\footnotesize
```{r, warning=F, error=F, eval=F}
for (i in seq_along(state_names)) {
```

---

# Poll prediction for each state {.t}

\footnotesize
```{r, warning=F, error=F, eval=F}
for (i in seq_along(state_names)) {
  ## subset the ith state
  state.data <- polls16 %>%
    filter(state == state_names[i])

```

---

# Poll prediction for each state {.t}

\footnotesize
```{r, warning=F, error=F, eval=F}
for (i in seq_along(state_names)) {
  ## subset the ith state
  state.data <- polls16 %>%
    filter(state == state_names[i])
  
  ## pull out the closest date (minimum days to election)
  min_days <- min(state.data$days_to_election)
```

---

# Poll prediction for each state {.t}

\footnotesize
```{r, warning=F, error=F, eval=F}
for (i in seq_along(state_names)) {
  ## subset the ith state
  state.data <- polls16 %>%
    filter(state == state_names[i])
  
  ## pull out the closest date (minimum days to election)
  min_days <- min(state.data$days_to_election)
  
  ## subset only the latest polls within the state
  state.data <- state.data %>%
    filter(days_to_election == min_days)
  
```

---

# Poll prediction for each state {.t}

\footnotesize
```{r, warning=F, error=F}
for (i in seq_along(state_names)) {
  ## subset the ith state
  state.data <- polls16 %>%
    filter(state == state_names[i])
  
  ## pull out the closest date (minimum days to election)
  min_days <- min(state.data$days_to_election)
  
  ## subset only the latest polls within the state
  state.data <- state.data %>%
    filter(days_to_election == min_days)
  
  ## compute the mean of the latest polls and store it
  poll.pred[i] <- mean(state.data$margin)
}
head(poll.pred)
```

---

# Poll prediction for each state (my way)

\footnotesize
```{r, results='hide'}
poll.list <- list()
state_names <- unique(polls16$state)

for (i in seq_along(state_names)) {
  state.data <- polls16 %>%
    filter(state == state_names[i]) %>%
    filter(days_to_election == min(days_to_election)) %>% 
    mutate(margin_poll = mean(margin)) %>%
    select(state, margin_poll)
  poll.list[[i]] <- state.data
  print(i)
}
PollPred <- do.call(rbind,poll.list)
head(PollPred)
```

---

# Comparing polls to outcomes

```{r, echo=FALSE, fig.width=4, fig.height=3}
plot <- full_join(PollPred, pres16, 
                   by = c("state" = "state_abb"))

plot %>%
  ggplot(aes(x = margin_poll, y = margin, label=state)) +
  geom_text() + 
  geom_abline(intercept = 0, slope = 1, linetype="dashed") +
  geom_hline(yintercept=0) +
  geom_vline(xintercept=0) +
  labs(y = "Actual Election Results",
       x = "Poll Results") +
  theme_bw()

```

