---
title: "PSC4375: Missing Data"
subtitle: "Week 3: Lecture 6"
author: "Prof. Weldzius"
institute: "Villanova University"
date: "Slides Updated: `r Sys.Date()`"
output:
  beamer_presentation:
    theme: "CambridgeUS"
    colortheme: "seahorse"
    fonttheme: "structurebold"
---

```{css,echo = F}
.small .remark-code { /*Change made here*/
  font-size: 85% !important;
}
.tiny .remark-code { /*Change made here*/
  font-size: 50% !important;
}
```

```{r,include=F}
set.seed(123)
options(width=60)
knitr::opts_chunk$set(fig.align='center',fig.width=9,fig.height=5,message=F,warning=F)
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
```

# Civilian attitudes and war against insurgency

  - War in Afghanistan: counter-insurgency war
    - Military against insurgents
    - Key to victory: winning hearts and minds of civilians
    - Aid provision, information campaign, minimizing civilian casualties
    
  - How does exposure to violence affect support for Taliban coalition?

\begin{center}
\includegraphics[width=.5\textwidth]{figs/afghan_map.png}
\end{center}
---

# Afghan study

```{r}
library(tidyverse)
data(afghan, package = "qss")
head(afghan[,1:8])
```

---

# Missing data

  - **Nonresponse**: respondent can't or won't answer question
    -Sensitive questions $\rightsquigarrow$ **social desirability bias**
    -Some countries lack official statistics like unemployment
    - Leads to missing data
    
  - Missing data in R: a special value `{\color{green}NA}`
  - Causes problems with calculating statistics:
  
```{r}
## prop. of those who got hurt by ISAF
mean(afghan$violent.exp.ISAF)
```
  
    
---

# Handling missing data in R (UPDATE TIDY)

  - Adding `na.rm = TRUE` to some functions removes missing data

```{r}
afghan %>% summarize(mean(violent.exp.ISAF, na.rm = TRUE))
```

  - Or, you can explicitly remove missing values using `na.omit()` function:
  
```{r}
afghan %>% summarize(mean(na.omit(violent.exp.ISAF)))
```
  
  - See number of `NA`s with `count()` + `group_by()`
  
```{r}
afghan %>%
  group_by(violent.exp.ISAF) %>%
  count()
summary(afghan$violent.exp.ISAF)
```
  
---

# Available-case vs. complete-case analysis

  - **Available-case analysis**: use the data you have for that variable:
  
```{r}
afghan %>%
  summarize(sum(!is.na(violent.exp.ISAF)))
afghan %>%
  summarize(mean(violent.exp.ISAF,na.rm=TRUE))
```

---

# Available-case vs. complete-case analysis

  - **Complete-case analysis**: only use units that have data on all variables 
    - Also called **listwise deletion**
    
```{r}
dim(na.omit(afghan))
afghan %>%
  na.omit() %>%
  summarize(mean(violent.exp.ISAF))
```
    
---

# Non-response and other biases

  - Nonresponse can create bias
  - More violent areas $\rightsquigarrow$ more non-response:
  
```{r}
afghan %>%
  group_by(province) %>%
  summarize(violent.exp.taliban = mean(is.na(violent.exp.taliban)),
            violent.exp.ISAF = mean(is.na(violent.exp.ISAF)))
```

  - $\rightsquigarrow$ oversampling citizens with less exposure to violence! 
