---
title: "Currency Wars in Retreat"
subtitle: "How Global Value Chains Affect Exchange Rate Politics"
author: "Prof. Weldzius"
institute: "Villanova University"
date: "Slides Updated: `r Sys.Date()`"
output:
  beamer_presentation:
    theme: "CambridgeUS"
    colortheme: "seahorse"
    fonttheme: "structurebold"
header-includes:
  - \usepackage{tikz}
  - \usepackage{graphicx} 
  - \usepackage{amssymb}
  - \usepackage{fontenc}
  - \usepackage{xcolor}    
  - \usepackage{array}
# classoption: "handout"  
---

```{css,echo = F}
.small .remark-code { /*Change made here*/
  font-size: 85% !important;
}
.tiny .remark-code { /*Change made here*/
  font-size: 50% !important;
}
```

```{r,include=F}
set.seed(123)
options(width=60)
knitr::opts_chunk$set(fig.align='center',fig.width=9,fig.height=5,message=F,warning=F)
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
library(tidyverse)
library(tinytex)
library(gridExtra)
```

# Are Currency Wars in Retreat?

- Exchange rate undervaluation worked for decades \pause
  - Japan (1970-1985) \pause
  - Germany (pre-EMU) \pause
  - China (1995-2005) \pause
  - Taiwan, South Korea, Israel, Sweden, etc. \pause
- By 2018, this practice has stopped; why?   \pause
- Argument: global value chains undercut the benefits of this strategy


---

# Global Value Chains over time

```{r, echo=FALSE, fig.height=3, fig.width=4}
# echo = FALSE stops Rmarkdown from showing anything in this chunk when you knit 
# fig.height and fig.width might need to be adjusted

df <- read_csv("~/Downloads/currencymanip2020.csv")

df %>%
  mutate(gvc_exp = exp(l_gvc_exp)) %>%
  filter(gvc_exp<15) %>%
  group_by(year) %>%
  mutate(gvc_exp_mean = mean(gvc_exp)) %>%
  ggplot(aes(x=year, y=gvc_exp, group=country_cd)) +
  geom_line(alpha=0) +
  geom_line(aes(x=year, y=gvc_exp_mean), size=1, alpha=0) +
  labs(x = "", y = "Global value chains/exports (%)") +
  scale_x_continuous(limits=c(1990,2020), breaks=seq(1990,2020,5)) +
  theme_bw()
```

---

# Global Value Chains over time

```{r, echo=FALSE, fig.height=3, fig.width=4}
# echo = FALSE stops Rmarkdown from showing anything in this chunk when you knit 
# fig.height and fig.width might need to be adjusted
df %>%
  mutate(gvc_exp = exp(l_gvc_exp)) %>%
  filter(gvc_exp<15) %>%
  group_by(year) %>%
  mutate(gvc_exp_mean = mean(gvc_exp)) %>%
  ggplot(aes(x=year, y=gvc_exp, group=country_cd)) +
  geom_line(alpha=.25) +
  geom_line(aes(x=year, y=gvc_exp_mean), size=1, alpha=0) +
  labs(x = "", y = "Global value chains/exports (%)") +
  scale_x_continuous(limits=c(1990,2020), breaks=seq(1990,2020,5)) +
  theme_bw()
```

---

# Global Value Chains over time

```{r, echo=FALSE, fig.height=3, fig.width=4}
# echo = FALSE stops Rmarkdown from showing anything in this chunk when you knit 
# fig.height and fig.width might need to be adjusted
df %>%
  mutate(gvc_exp = exp(l_gvc_exp)) %>%
  filter(gvc_exp<15) %>%
  group_by(year) %>%
  mutate(gvc_exp_mean = mean(gvc_exp)) %>%
  ggplot(aes(x=year, y=gvc_exp, group=country_cd)) +
  geom_line(alpha=.25) +
  geom_line(aes(x=year, y=gvc_exp_mean), size=1, color="red") +
  labs(x = "", y = "Global value chains/exports (%)") +
  scale_x_continuous(limits=c(1990,2020), breaks=seq(1990,2020,5)) +
  theme_bw()
```

---

# What do I expect?

- Hypothesis 1: the more dependent on trade, the more likely a country is to devalue their exchange rate \pause

- Hypothesis 2: the more dependent on trade, conditional on your global value chain dependence, the **less** likely to devalue \pause 

\begin{align*}
\text{devaluation}_{it} = & \alpha_{it} + \beta_1 (\text{export dependence})_{it} + \beta_2 (\text{GVC dependence})_{it} + \\
& \beta_3 (\text{export dependence})_{it}\times(\text{GVC dependence})_{it} + \epsilon_{it}
\end{align*}

- $i$ indicates country
- $t$ indicates year

---

# Does GVC dependence impact exchange rate policy?

```{r message=FALSE, include=FALSE} 
# Here you can include your statistical model. You'll want to output the results in a readable way. 
# Use this link to help with formatting these tables (see "regression table"); note that you'll have to use some new libraries (and note what they include at the beginning of the R chunk, e.g., "message = FALSE"): https://zief0002.github.io/book-8252/pretty-printing-tables-in-markdown.html
library(lfe)
df$country_factor <- as.factor(df$country_cd)
df$year_factor <- as.factor(df$year)

m1 <- felm(l_forex_gdp_f1 ~ l_exp_goods_gdp | year_factor + country_factor | 0 | country_factor, data=df)

m2 <- felm(l_forex_gdp_f1 ~ l_exp_goods_gdp*l_gvc_exp | year_factor + country_factor | 0 | country_factor, data=df)
```

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis'} 
library(stargazer)
table <- stargazer(m1, m2, 
          type = "latex",
          header = FALSE,
          no.space = TRUE,
          float = FALSE,
          keep.stat = c("n","adj.rsq"), # only keep what I need
          column.labels = c("Model 1", "Model 2"),
          colnames = FALSE,
          model.numbers = FALSE,
          dep.var.caption = " ",
          dep.var.labels = "",
          covariate.labels = c("Exports/GDP", "GVC/Exports","(Exp/GDP)x(GVC/Exp)"),
          notes.align = "l"
)

#table <- table[-c(3:7)] # removes the top lines of the Stargazer table so we can see more rows
cat(paste(table, collapse = "\n"))
```

---

# Conclusion and next steps

- The data supports my hypotheses: export dependence *does* lead to currency undervaluation, but this relationship is moderated by global value chain dependence \pause 
- Next steps: \pause

  1. What control variables should I include? \pause
  2. I used a linear model, but could there be a non-linear relationship?

---

# Appendix 1

```{r, echo=FALSE, fig.height=3, fig.width=4}
# echo = FALSE stops Rmarkdown from showing anything in this chunk when you knit 
# fig.height and fig.width might need to be adjusted

df <- read_csv("~/Downloads/currencymanip2020.csv")

df %>%
  mutate(gvc_exp = exp(l_gvc_exp)) %>%
  ggplot(aes(x=l_gvc_exp)) +
  geom_density()
```